---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# designit

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of designit is to ...

## skeleton

* a metric of how balanced?
* check ^2 and ^3, it should be different (3 better)

### algorithm

- generate formula
  - .^m m = number of factors
  - check model.matrix
  - check how many to remove, and what order
  - if one order is not complete, then do random draw from that?
  - or use optMonteCarlo <-- this doesn't work
- optFederov
- optBlock
- rename Block and levels
- do it again

**should we support more than 2 levels?**

**what to do with numeric covariates**


```{r}
a <- readxl::read_excel('inst/extdata/CEA-PRIT_CEAmodels_for randomization.xlsx')
```


## Installation

TODO

## Example
```{r message=FALSE}
library(designit)
# define samples data.frame
samples <- data.frame(a='a', b=letters[1:3], c=c('b', 'b', 'c'))
samples

bc <- BatchContainer$new(
  dimensions = list(
    "plate" = 3,
    "row" = list(values = letters[1:3]),
    "column" = list(values = c(1, 3))
  ),
  exclude = data.frame(plate = 1, row = "a", column = c(1, 3), stringsAsFactors = F)
)

bc

bc$exclude
bc$n_locations
bc$n_excluded
bc$n_available
bc$locations_df

bc$distribute_samples(samples, random_seed=1)

head(bc$get_samples())
bc$get_samples(remove_empty_locations=TRUE)

# You can reassign samples starting from the current assignment.
bc$distribute_samples(random_seed=2)
head(bc$get_samples())

# Results should be reproducible if the seed is set.
bc$distribute_samples(random_seed=1)
bc$get_samples(remove_empty_locations=TRUE)
```

```{r}
library(tidyverse)
x <- expand.grid(b = paste0('b', 1:4),
                 a = LETTERS[1:4]) %>% 
  slice(-3:-4)
```


Example dataset:

Cell: 5 levels (A, B, C, D, E)
Trt: 2 levels (N, T)
Time: 2 levels (24, 48)

2 batches:
Extract: 5 levels (1-5)
Prep: 2 levels (1-2)

all combinations, 80 expt
