---
output:
  github_document:
    fig_width: 5
    fig_height: 5
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# designit <a href="https://bedapub.github.com/designit/"><img src="man/figures/logo.svg" align="right" height="139" /></a>

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Documentation](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://bedapub.github.io/designit/)
<!-- badges: end -->

The goal of designit is to generate optimal sample allocations for experimental designs. 

## Installation

You can install the development version from GitHub with:

```{r echo=TRUE, eval=FALSE}
# install.packages("devtools")
devtools::install_github("BEDApub/designit")
```

## Usage

The main class used is `BatchContainer`, which holds the dimensions for sample allocation.
After creating such a container, a list of samples can be allocated in it using a given assignment function.

### Creating a table with sample information

```{r "samples table", message=FALSE}
library(tidyverse)
library(designit)

data("longitudinal_subject_samples")

# we use a subset of longitudinal_subject_samples data
subject_data <- longitudinal_subject_samples %>% 
  filter(Group %in% 1:5, Week %in% c(1,4)) %>% 
  select(SampleID, SubjectID, Group, Sex, Week) %>%
  # with two observations per patient
  group_by(SubjectID) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  select(SubjectID, Group, Sex) %>%
  distinct()

head(subject_data)
```

### Creating a `BatchContainer` and assigning samples
```{r "random assignment"}
# a batch container with 3 batches and 11 locations per batch
bc <- BatchContainer$new(
  dimensions = list("batch" = 3, "location" = 11),
)

# assign samples randomly
set.seed(17)
assign_random(bc, subject_data)

bc$get_samples() %>%
  ggplot() +
  aes(x = batch, fill = Group) +
  geom_bar()
```

Random assignemt of samples to batches produced an uneven distribution.

### Optimizing the assignemnt
```{r "optimized assignment", warning=FALSE}
# set scoring functions
bc$scoring_f <- list(
  # first priority, groups are evenly distributed
  group = osat_score_generator(batch_vars = "batch", 
                               feature_vars = "Group"),
  # second priority, sexes are evenly distributed
  sex = osat_score_generator(batch_vars = "batch", 
                             feature_vars = "Sex")
)

trace <- optimize_design(bc, max_iter = 150, quiet = TRUE)

bc$get_samples() %>%
  ggplot() +
  aes(x = batch, fill = Group) +
  geom_bar()

# show optimization trace
plot(trace)
```

## Examples

See vignettes `vignette("basic_examples")`.

## Acknowledgement

The logo is inspired by [DALL-E 2](https://openai.com/dall-e-2/) and
[pipette icon by gsagri04](https://openclipart.org/detail/140941/micropipette).
