---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# designit

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of designit is to generate optimal sample allocations for experimental designs. Documentation is available [here](http://bioinfo.bas.roche.com:8080/biomics/_doc/packages/designit/).

## Installation

You can install the development version from Roche GitHub with:

```{r echo=TRUE, eval=FALSE}
# install.packages("devtools")
devtools::install_gitlab("BEDA/designit", host = "code.roche.com")
```

## Usage

The main class used is `BatchContainer`, which holds the dimensions for sample allocation.
After creating such a container, a list of samples can be allocated in it using a given assignment function.

```{r message=FALSE}
library(designit)
# define samples data.frame
samples <- data.frame(a='a', b=letters[1:3], c=c('b', 'b', 'c'))
samples

bc <- BatchContainer$new(
  dimensions = list(
    "plate" = 3,
    "row" = list(values = letters[1:3]),
    "column" = list(values = c(1, 3))
  ),
  exclude = data.frame(plate = 1, row = "a", column = c(1, 3), stringsAsFactors = F)
)

bc

bc$exclude
bc$n_locations
bc$n_excluded
bc$n_available
bc$locations_df

set.seed(1)
assign_random(bc, samples)

head(bc$get_samples())
bc$get_samples(remove_empty_locations=TRUE)

# You can reassign samples starting from the current assignment.
set.seed(2)
assign_random(bc)
head(bc$get_samples())

# Results should be reproducible if the seed is set.
set.seed(1)
assign_random(bc)
bc$get_samples(remove_empty_locations=TRUE)
```

## Examples

See vignettes `vignette("basic_examples")`.
