---
title: "In vivo examples computed"
output: html_fragment
knit: (\(input, ...) rmarkdown::render(input, output_dir = "vignettes"))
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  fig.width = 6,
  fig.height = 6
)
```
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(designit)
library(dplyr)
library(forcats)
library(assertthat)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
library(cowplot)
```

```{r include=FALSE}
# load all data and functions from the main vignette
.precompile <- TRUE
knitr::purl("../invivo_study_design.Rmd",
            output = tempfile(),
            documentation = 0) %>% 
  source(echo = FALSE)
```

# Calculating the design

A full run of the three wrapper functions is executed below, printing messages
on the level of processing steps, but not the iterations within every
optimization.

```{r echo=TRUE}
set.seed(44)

# Assign treatments to animals, respecting user provided as well as passed constraints
design_trt <- InVivo_assignTreatments(invivo_study_samples, invivo_study_treatments,
  form_cages_by = c("Strain", "Sex", "Litter_combine_females"),
  balance_treatment_vars = c("Strain", "Sex", "ArrivalWeight", "AgeGroup"),
  n_shuffle = c(rep(5, 200), rep(3, 300), rep(2, 500), rep(1, 3000)),
  quiet_process = FALSE,
  quiet_optimize = TRUE
)

# Form cages with reasonable animal numbers and compliant with all constraints
design_cage <- Invivo_assignCages(design_trt,
  cagegroup_vars = c("Treatment", "Strain", "Sex", "Litter_combine_females"),
  unique_vars = c("Earmark"),
  balance_cage_vars = c("ArrivalWeight", "AgeGroup"),
  n_min = 2, n_max = 5, n_ideal = 2, prefer_big_groups = T, strict = F,
  maxiter = 1000,
  quiet_process = FALSE,
  quiet_optimize = TRUE
)

# Arrange cages in sub-grid of one rack (or several racks), avoiding spatial clusters
design_rack <- Invivo_arrangeCages(design_cage,
  distribute_cagerack_vars = c("Treatment", "Strain", "Sex"),
  rack_size_x = 7,
  rack_size_y = 10,
  n_shuffle = c(rep(5, 100), rep(3, 200), rep(2, 300), rep(1, 500)),
  quiet_process = FALSE,
  quiet_optimize = TRUE
)
```

# Visualization of the study design

## Cage composition

There are `r dplyr::n_distinct(design_rack$CageNr)` cages in total.

Strains and age groups should be evenly split (balanced) across the treatments. Also,in each cage there should be
only animals with the same treatment, strain and sex.

Females are exclusively used for treatment 2, as was specified in the treatment list.

```{r echo=FALSE, fig.width = 8, fig.height = 5, warning=FALSE, message=FALSE}

p1 <- ggplot2::ggplot(design_rack, aes(x = factor(CageNr), y = Strain, color = Treatment)) +
  facet_grid(~Treatment, scales = "free_x", space = "free_x") +
  geom_jitter(width = 0.2, height = 0) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  labs(x = "Cage Nr", y = "", title = "Distribution of strains")

p2 <- ggplot2::ggplot(design_rack, aes(x = factor(CageNr), y = Sex, color = Treatment)) +
  facet_grid(~Treatment, scales = "free_x", space = "free_x") +
  geom_jitter(width = 0.2, height = 0) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  labs(x = "Cage Nr", y = "", title = "Distribution of sex")

p3 <- ggplot2::ggplot(design_rack, aes(x = factor(CageNr), y = BirthDate, color = Treatment)) +
  facet_grid(~Treatment, scales = "free_x", space = "free_x") +
  geom_jitter(width = 0.2, height = 0) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  labs(x = "Cage Nr", y = "", title = "Distribution of birth dates")

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```

## Body weights

Body weights should be balanced across treatments as well as possible.

The plot illustrates that this is true for the overall weight distribution (box plots). Interestingly, as there
are females (associated with considerable less body weight) involved in treatment 2, the optimization favored the
selection of heavier males in this group to compensate, achieving better cross-treatment balance of this factor.

Red diamonds mark the mean values for a specific sex within each treatment group.

```{r echo=FALSE, fig.width = 8, fig.height = 5, warning=FALSE, message=FALSE}

ggplot2::ggplot(design_rack, aes(x = Sex, y = ArrivalWeight, color = Treatment, fill = Treatment)) +
  facet_grid(Strain ~ Treatment, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  geom_boxplot(aes(group = Treatment), alpha = 0.2) +
  geom_jitter(width = 0.2, height = 0.2) +
  stat_summary(fun = "mean", colour = "red", shape = 18, size = 10, geom = "point", aes(group = Treatment), alpha = 0.3, show.legend = F) +
  labs(y = "Arrival weight [g]")
```

## Cage arrangement in rack

The following plots show the organization of the cage rack, individual cages colored by different
variables each time.

```{r echo=FALSE, fig.width = 10, fig.height = 4, warning=FALSE, message=FALSE}

p1 <- Invivo_plotByRack(design_rack, colorBy = "Treatment", showLegend = T)

p2 <- Invivo_plotByRack(design_rack, colorBy = "Strain", showLegend = T)

p3 <- Invivo_plotByRack(design_rack, colorBy = "Sex", showLegend = T)

cowplot::plot_grid(plotlist = c(p1, p2, p3), ncol = 3)
```


## Individual animals in cages

Finally, an overview plot illustrates the placement of animals in the cages. Notice the distinct earmarks
within each cage, a 'soft' design constraint that could be achieved with the given solution.

```{r echo=FALSE, fig.width = 7, fig.height = 5, warning=FALSE, message=FALSE, results="asis"}

pl <- Invivo_plotByRack(design_rack, colorBy = "Treatment", showAnimals = TRUE)

cowplot::plot_grid(plotlist = pl)
```
