---
title: "Plate scoring example 2"
output: html_fragment
knit: (\(input, ...) rmarkdown::render(input, output_dir = "vignettes/cached"))
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(designit)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Example 2: Scoring two plates at once

(latin square for each plate should give the best score)

We set up in one go 2 plate scoring functions, each one acting locally on a
specific plate, plus one osat score to guarantee uniform distribution of groups
across plates.

The initial sample allocation (by assign_in_order) leads to a poor starting
point since each plate has only 2 of the 4 groups represented.

This is not a problem as long as we make sure that initial permutations are
likely to remedy the situation. That's why we ensure 10 pairwise sample swaps
for the first iterations.

```{r}
# Setting up the batch container
example2 <- BatchContainer$new(
  dimensions = c(
    plate = 2,
    row = 4, col = 4
  )
)

# Add samples to container
assign_in_order(example2, samples = tibble::tibble(
  Group = c(rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 8)),
  ID = 1:32
))

bc <- example2$copy()

bc$scoring_f <- c(mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group"),
  osat_plate = osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
)

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: Initial sample arrangement"
)

bc$score()
```

```{r}
set.seed(41)
trace <- optimize_design(bc,
  n_shuffle = c(rep(10, 10), rep(3, 90), rep(2, 100), rep(1, 1400)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
  aggregate_scores_func = worst_score,
  quiet = TRUE
)
```

```{r}
trace$elapsed

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: Design created by swapping samples 'globally' across the plates"
)
```

While this 'global' optimization is possible, it does probably not converge to
an (almost) ideal solution in an acceptable time if there are more samples
involved. This is due to a lot of unproductive sample swapping happening across
the plates.

One way to address this: we may split the optimization into two cycles, first
assigning samples to plates (balancing groups), then improving the positions of
the samples within each plate. This motivates the use of a dedicated sample
permutation function which takes the plate structure into account and only
shuffles samples around within one plate.

```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))

set.seed(42)
optimize_design(bc,
  quiet = TRUE,
  max_iter = 200, # this is set to shorten vignette run-time, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: 'Plate wise' design\nStep 1: after allocating samples to plates"
)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

bc$score()
```

```{r}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 400,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate")),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm,
  quiet = TRUE
)
```

```{r}
trace

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: 'Plate wise' design\nStep 2: after arranging samples within plates"
)
```

In this case, the shuffling function exchanges 1 pair of sample assignments
every time (the default). However, any number of constant swaps or a swapping
protocol (formally a vector of integers) can be supplied as well.


Now for the most efficient solution: we start again by first assigning samples
to plates (balancing groups), then making use of the independence of the two
within-plate optimizations and improving them one after the other.

This is possible by passing the argument \code{restrain_on_subgroup_levels} to
the function that generates the permutations. It enforces permutation only to
happen first within plate 1, then within plate 2, so that the two scores can be
optimized in succeeding runs.

```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
```

```{r}
set.seed(42)
trace <- optimize_design(bc,
  quiet = TRUE,
  max_iter = 150, # this is set to shorten vignette run-time, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)
```

```{r}
trace

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: 'Serial plate' design\nStep 1: after allocating samples to plates"
)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

bc$score()
```

```{r}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 150,
  quiet = TRUE,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate"), restrain_on_subgroup_levels = c(1)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r}
trace

bc$score()
```

```{r}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 550,
  quiet = TRUE,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate"), restrain_on_subgroup_levels = c(2)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r}
trace

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex2: 'Serial plate' design\nStep 2: after optimizing each plate in turn"
)
```
