---
title: "Plate scoring example 1"
output: html_fragment
knit: (\(input, ...) rmarkdown::render(input, output_dir = "vignettes"))
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(designit)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Example 1: An expensive way to construct a 4x4 latin square (one plate)

(latin square should give the best score)

First using a combination of two OSAT scores (for row and column).

This usually produces a latin square when using the squared L2 norm (L2s) for
aggregation of the 2 scores.

```{r}
# Setting up the batch container
example1 <- BatchContainer$new(
  dimensions = c(
    plate = 1,
    row = 4, col = 4
  )
)

# Add samples to container
# Need unique Sample ID. Can we drop this constraint?
assign_in_order(example1,
  samples = tibble::tibble(
    Group = rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 4),
    ID = 1:16
  )
)

# The following does not work (an gives a constant score of 144!)
# example1$scoring_f <- osat_score_generator(batch_vars = c("row","col"), feature_vars = c("Group"))
# First analysis of problem indicates that osat_score generates a full row*col vector of 'ideal scores'
# which are in fact the same value, implying an identical overall result as each position can be either
# allocated by 1 sample or 0 samples, the sum of 1's being the sample count.
# --> don't use osat_score if there's a lack of samples as compared to possible positioning

bc <- example1$copy()

# # Set scoring function
bc$scoring_f <- list(
  Row.Score = osat_score_generator(batch_vars = c("row"), feature_vars = c("Group")),
  Column.Score = osat_score_generator(batch_vars = c("col"), feature_vars = c("Group"))
)
```

```{r}

set.seed(41)

trace <- optimize_design(bc,
  max_iter = 300, # this is set to shorten vignette run-time based on known random seed, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r}
trace

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex1: Using OSAT scores for plate design\n(not the recommended way!)"
)
```

Now using a dedicated scoring for the group distances on a plate. 

This should reliably lead to a nice symmetry-bearing latin square design with
only a one-dimensional score to look at.

```{r}
bc <- example1$copy()

bc$scoring_f <- mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group")
```

```{r}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 1000, # this is set to shorten vignette run-time based on random seed, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  quiet = TRUE
)
```

```{r}
trace$elapsed

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Ex1: Using a dedicated plate scoring function:\nThis should show a latin square!"
)
```
