---
title: "Basic examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(designit)
library(tidyverse)
```

## Example 1

This is a basic example:

```{r example1}
anno <- tribble(
  ~Cell,         ~Trt, ~Type, ~Rep,
   "WT",    "treated",   "A",    1,
   "WT",    "treated",   "A",    2,  
   "WT", "nontreated",   "A",    1,
   "WT", "nontreated",   "A",    2,
   "WT",    "control",   "A",    1,
   "WT",    "control",   "A",    2,
   "WT",    "treated",   "B",    1,
   "WT",    "treated",   "B",    2,  
   "WT", "nontreated",   "B",    1,
   "WT", "nontreated",   "B",    2,
   "WT",    "control",   "A",    1,
   "WT",    "control",   "A",    2,
   "KO",    "treated",   "A",    1,
   "KO",    "treated",   "A",    2,  
   "KO", "nontreated",   "A",    1,
   "KO", "nontreated",   "A",    2,
   "KO",    "treated",   "B",    1,
   "KO",    "treated",   "B",    2,  
   "KO", "nontreated",   "B",    1,
   "KO", "nontreated",   "B",    2,  
)
```


## Plate layout with two factors

Samples of a 2-condition invivo experiment are to be
placed on 48 well plates.

These are the conditions

```{r}

# conditions to use
conditions <- data.frame(group = c(1, 2, 3, 4, 5),
                      treatment = c("vehicle", "TRT1", "TRT2",
                                    "TRT1", "TRT2"),
                      dose = c(0, 25, 25, 50, 50))

knitr::kable(conditions)
```

We will have 3 animals per groups with 4 replicates each

```{r}
# sample table (2 animals per group with 3 replicates)
n_reps <- 4
n_animals <- 3
samples <- dplyr::bind_rows(replicate(n_animals, conditions, simplify = FALSE), .id = 'animal') 
samples <- dplyr::bind_rows(replicate(n_reps, samples, simplify = FALSE), .id = 'replicate') %>%
  dplyr::mutate(SampleID = paste0(treatment,'_',animal,'_',replicate),
                AnimalID = paste0(treatment,'_',animal))

DT::datatable(samples)
```

Corner wells of the plates should be left empty.
This means on a 48 well plate we can place 44 samples.
Since we have `r nrow(samples)` samples, they will fit on
`r ceiling(nrow(samples)/44)` plates

```{r}
n_samp <- nrow(samples)
n_loc_per_plate <- 48 - 4
n_plates <- ceiling(n_samp / n_loc_per_plate)

exclude_wells <- expand.grid(plate = seq(n_plates), column = c(1,8), row = c(1,6))
```

Create a BatchContainer object that provides all possible locations

```{r}
bc <- BatchContainer$new(
  dimensions = c('plate' = n_plates, 'column' = 8, 'row' = 6),
  exclude = exclude_wells
)
bc

bc$n_available
bc$exclude
```

Use random assignment function to place samples to plate locations

```{r}
bc$assign_samples(
  samples = samples,
  assignment_function = assign_random,
  random_seed = 42
)

bc$get_samples()
bc$get_samples(remove_empty_locations = TRUE)

```

Plot of the result

```{r}
library(ggplot2)
bc$get_samples(remove_empty_locations = TRUE) %>% 
  ggplot(aes(x = factor(column), y = factor(-row), fill = treatment, alpha = factor(dose))) + 
  geom_tile() +
  facet_wrap(~plate) + 
  theme_bw() 
  
```

