---
title: "Basic example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(designit)
library(ggplot2)
library(dplyr)
library(tidyr)

```

# Plate layout with two factors

## The samples

Samples of a 2-condition in-vivo experiment are to be
placed on 48 well plates.

These are the conditions

```{r}

# conditions to use
conditions <- data.frame(group = c(1, 2, 3, 4, 5),
                      treatment = c("vehicle", "TRT1", "TRT2",
                                    "TRT1", "TRT2"),
                      dose = c(0, 25, 25, 50, 50))

gt::gt(conditions)
```

We will have 3 animals per groups with 4 replicates each

```{r}
# sample table (2 animals per group with 3 replicates)
n_reps <- 4
n_animals <- 3
animals <- dplyr::bind_rows(replicate(n_animals, conditions, simplify = FALSE), 
                            .id = 'animal') 
samples <- dplyr::bind_rows(replicate(n_reps, animals, simplify = FALSE), 
                            .id = 'replicate') %>%
  dplyr::mutate(SampleID = paste0(treatment,'_',animal,'_',replicate),
                AnimalID = paste0(treatment,'_',animal)) 

samples %>% head(10) %>% gt::gt()
```
## Plate layout requirements

Corner wells of the plates should be left empty.
This means on a 48 well plate we can place 44 samples.
Since we have `r nrow(samples)` samples, they will fit on
`r ceiling(nrow(samples)/44)` plates

```{r}
n_samp <- nrow(samples)
n_loc_per_plate <- 48 - 4
n_plates <- ceiling(n_samp / n_loc_per_plate)

exclude_wells <- expand.grid(plate = seq(n_plates), column = c(1,8), row = c(1,6))
```

## Setting up a Batch container

Create a BatchContainer object that provides all possible locations

```{r}
bc <- BatchContainer$new(
  dimensions = c('plate' = n_plates, 'column' = 8, 'row' = 6),
  exclude = exclude_wells
)
bc

bc$n_available
bc$exclude
bc$get_locations() %>% head()
```

## Moving samples

Use random assignment function to place samples to plate locations

```{r}
assign_random(bc, samples)

bc$get_samples()
bc$get_samples(remove_empty_locations = TRUE)

```


Plot of the result using the `plot_plate` function

```{r, fig.width=6, fig.height=4}
plot_plate(
  bc$get_samples(remove_empty_locations = FALSE),
           plate = plate, column = column, row = row,
           .color = treatment, .alpha = dose) 
```

To move individual samples or manually assigning all locations we can use the 
`batchContainer$move_samples()` method

To swap two or more samples use
```{r, fig.width=6, fig.height=3.5}
bc$move_samples(src = c(1L, 2L), dst = c(2L, 1L))

plot_plate(bc$get_samples(remove_empty_locations = TRUE), plate = plate, column = column, row = row,
           .color = treatment, .alpha = dose)
```

To assign all samples in one go, use the option `location_assignment`.
The example below orders samples by ID and adds the empty locations afterwards
```{r, fig.width=6, fig.height=3.5}
bc$move_samples(location_assignment = as.integer(c(1:nrow(samples), rep(NA, (bc$n_available - nrow(samples))))))

plot_plate(bc$get_samples(remove_empty_locations = TRUE, include_id = TRUE), 
           plate = plate, column = column, row = row,
           .color = .sample_id)
```

## Scoring a layout

To evaluate how good a layout is, we need a scoring function. 
This we also assign to the batch container.

This function will assess how well treatment 
and dose are balanced across the two plates.

```{r}
bc$scoring_f <- osat_score_generator(batch_vars = "plate", 
                                     feature_vars = c("treatment", "dose"))
```

## Run an optimization

The optimization procedure is invoked with e.g. `assign_score_optimize_shuffle`.
Here we use a simple shuffling schedule: 
swap 10 samples for 100 times, then swap 2 samples for 400 times.

```{r}
trace <- assign_score_optimize_shuffle(bc,
                              # shuffling schedule
                              n_shuffle=c(rep(10, 200), rep(2, 400))
)
```

Development of the score can be viewed with

```{r, fig.width=3.5, fig.height=3}
trace$plot()
```


The layout after plate batching looks the following

```{r, fig.width=6, fig.height=3.5}
plot_plate(bc$get_samples(remove_empty_locations = TRUE), 
           plate = plate, column = column, row = row,
           .color = treatment, .alpha = dose)
```

Looking at treatment, we see it's evenly distributed across the plates

```{r, fig.width=4, fig.height=3.5}
ggplot(bc$get_samples(remove_empty_locations = TRUE),
       aes(x = treatment, fill = treatment)) + geom_bar() +  
   facet_wrap(~plate)
```

## Customizing the plate layout

To properly distinguish between empty and excluded locations one can do the
following.

```{r}
color_vector <- c(TRT1 = "blue", TRT2 = "orange", vehicle = "purple")
plot_plate(
  bind_rows(bc$get_samples(remove_empty_locations = FALSE) %>% 
              mutate(treatment = ifelse(is.na(treatment), "empty", treatment)),
            bc$exclude),
           plate = plate, column = column, row = row,
           .color = treatment, .alpha = dose) +
  scale_fill_manual(values = c(color_vector, empty = 'white'), na.value="darkgray")
```

