---
title: "Cages"
author: "Balazs Banfai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 6,
  fig.width = 8
)
has_printr <- requireNamespace("printr", quietly = TRUE) # for printing help pages nicely

library(designit)
library(tidyverse)
```

# Description of data

```{r echo=FALSE}
if (has_printr) {
  help("cages", package = "designit", help_type = "text")
} else {
  help("cages", package = "designit", help_type = "text") %>%
    as.character() %>%
    utils:::.getHelpFile() %>%
    tools:::Rd2txt(outputEncoding = "UTF-8")
}
```

# Plots of cage layout

```{r, include=FALSE}
plot_cages <- function(data, var, colors, .row = CageRow, .col = CageCol) {
  varname <- as_label(enquo(var))
  data %>%
    distinct({{ .row }}, {{ .col }}, {{ var }}) %>%
    ggplot() +
    aes(x = 1, y = 1, fill = {{ var }}) +
    facet_grid(rows = vars({{ .row }}), cols = vars({{ .col }})) +
    geom_raster(alpha = 0.9, position = position_fill()) +
    scale_fill_manual(values = colors) +
    labs(title = paste0("Cage by ", varname)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}
```

```{r}
plot_cages(cages, Group, colors = c("darkorange", "cornflowerblue"))
```

```{r}
plot_cages(cages, Sex, colors = c("violet", "dodgerblue"))
```

```{r}
plot_cages(cages, Genotype, colors = c("orange", "gray"))
```

```{r}
plot_cages(cages, Group, colors = c("darkorange", "cornflowerblue")) +
  geom_text(aes(label = AnimalID),
    data = cages,
    position = position_fill(vjust = 0.5, reverse = TRUE)
  ) +
  labs(subtitle = "Animals by cages")
```

# Inspect solution

```{r}
ggplot(cages, aes(x = CageNr, y = Litter, color = Sex)) +
  facet_wrap(~Group, scales = "free_x", ncol = 4) +
  geom_jitter(width = 0.2, height = 0) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_color_manual(values = c("violet", "dodgerblue")) +
  geom_label(aes(label = Earmark), hjust = 0, size = 3, nudge_x = 0.2, nudge_y = 0.2)
```

```{r}
ggplot(cages, aes(x = CageNr, y = BirthDate, color = Sex)) +
  facet_wrap(~Group, scales = "free_x", ncol = 4) +
  geom_jitter(width = 0.2, height = 0.2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
ggplot(cages, aes(x = Group, y = ArrivalWeight, color = Sex, fill = Sex)) +
  facet_wrap(~Genotype, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_boxplot(aes(group = Group), alpha = 0.2) +
  geom_jitter(width = 0.2, height = 0.2) +
  stat_summary(
    fun = "mean", colour = "red",
    size = 10, geom = "point",
    aes(group = Genotype),
    shape = 18, alpha = 0.3, show.legend = F
  )
```

# Reproducing this solution

Goal: same number of animals get the same treatments (_Group1_ and _Group2_) for each `Genotype` and `Sex` based on the covariates (`Litter`, `Age`, `Weight`)

1) start with males (more constraints):
  * cages with constant treatment and males from one litter
  * as few isolated males as possible
    (2 animals per cage, but not more than 3-4)
  * don't mix sexes
  * take into account covariates as much as possible
2) females
  * distribute based on all covariates
  
Problems:
 
  * number of cages is not fixed
  * number of cages per treatment is also not fixed
  * earmarks have to be unique within each cage
  
