---
title: "Using custom shuffle schedule"
output: rmarkdown::html_vignette
params:
  iterations: 100
  n_samples: 50
vignette: >
  %\VignetteIndexEntry{Using custom shuffle schedule}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(designit)
devtools::load_all("..")
library(tidyverse)
```

In this example we would like to distribute animals among cages with constraints:

* There should be not more than one male per cage.
* Number of treatment/control animals should be comparable per cage
* Average weight per cage should be comparable between cages

```{r}
set.seed(43)
samples <- tibble(
  id = 1:params$n_samples,
  sex = sample(c("F", "M"), params$n_samples, replace = TRUE, prob = c(0.8, 0.2)),
  group = sample(c("treatment", "control"), params$n_samples, replace = TRUE),
  weight = runif(params$n_samples, 20, 30)
)
samples %>%
  head()
```

```{r}
samples %>%
  count(sex)
```
We create a `BatchContainer` with 11 cages and 5 positions per cage. Note that positions do not
actually matter; this is just to limit the number of animals per cage.

We start by assigning samples randomly.

```{r}
set.seed(42)
bc <- BatchContainer$new(
  dimensions = c("cage" = 11, "position" = 5)
)
assign_random(bc, samples)
bc
```

Functions to plot number of males per cage, weights per cage and treatment/control ratios.
```{r}
males_per_cage <- function(bc) {
  bc$get_samples() %>%
    filter(sex == "M") %>%
    count(cage) %>%
    ggplot(aes(cage, n)) +
    geom_col()
}

weight_d <- function(bc) {
  bc$get_samples() %>%
    ggplot(aes(factor(cage), weight)) +
    geom_violin() +
    geom_point() +
    stat_summary(fun = mean, geom = "point", size = 2, shape = 23, color = "red")
}

group_d <- function(bc) {
  bc$get_samples(remove_empty_locations = TRUE) %>%
    ggplot(aes(factor(cage), fill = group)) +
    geom_bar(position = "fill")
}
```

```{r}
males_per_cage(bc)
weight_d(bc)
group_d(bc)
```
This function can be used to provide shuffling schedule. `src` and `dst` are expressions which
are used to select first and second locations to exchange. `change_batch_variable` can be used
to enforce change of batch variable values between the two locations.

```{r}
shuffle_with_constraints <- function(src, dst = TRUE, change_batch_variable = NULL) {
  src <- enquo(src)
  dst <- enquo(dst)
  function(dt, i) {
    src_ind <- which(rlang::eval_tidy(src, dt))
    assertthat::assert_that(length(src_ind) > 0,
      msg = "source conditions not satisfied"
    )
    src_pos <- sample(src_ind, 1)
    # different from src
    dst_log <- seq(nrow(dt)) != src_pos
    dst_log <- dst_log & rlang::eval_tidy(dst, dt)
    if (!is.null(change_batch_variable)) {
      dst_log <- dst_log & dt[[change_batch_variable]] != dt[[src_pos, change_batch_variable]]
    }
    dst_ind <- which(dst_log)
    if (length(dst_ind) == 0) {
      warning("Cannot find destanation matching the constraints")
      list(src = NULL)
    } else {
      dst_pos <- sample(dst_ind, 1)
      list(
        src = c(src_pos, dst_pos),
        dst = c(dst_pos, src_pos)
      )
    }
  }
}
```

First, we use OSAT scoring function to ensure even distribution of males among cages. Only `cage`
and `sex` interactions are considered in the scoring function. We only use 10 iterations, since
shuffling is limited to locations with males and enforces change of cage on every iteration.

```{r}
bc$scoring_f <- function(samples) {
  osat_score(
    samples,
    "cage",
    "sex"
  )$score
}

set.seed(10)

res <- assign_score_optimize_shuffle(
  bc,
  shuffle_proposal = shuffle_with_constraints(
    sex == "M",
    change_batch_variable = "cage"
  ),
  iterations = 10
)

plot(res)
```
We expect the distribution of males become even, while other variables are not significantly
affected.

```{r}
males_per_cage(bc)
weight_d(bc)
group_d(bc)
```

Here we only define our custom scoring function which ensures even distribution of weights and
treatment/control groups. Only female samples are shuffled and male samples are kept in their
locations. We also ensure that on every iteration the cage number is changed; we do this because
`position` dimension does affect actual animal allocation.

```{r}
bc$scoring_f <- function(samples) {
  avg_w <- samples[, mean(weight, na.rm = TRUE)]
  avg_w_per_cage <- samples[!is.na(weight), mean(weight), by = cage]$V1
  trt_per_cage <- samples[!is.na(group), sum(group == "treatment") / .N, by = cage]$V1

  w_score <- mean((avg_w - avg_w_per_cage)**2)
  trt_score <- mean((trt_per_cage - 0.5)**2)
  w_score + 10 * trt_score
}

set.seed(12)
res <- assign_score_optimize_shuffle(bc,
  shuffle_proposal = shuffle_with_constraints(
    sex == "F",
    is.na(sex) | sex != "M",
    "cage"
  ),
  n_shuffle = c(rep(10, 20), rep(5, 20), rep(3, 20), rep(1, 140)),
  iterations = 200
)
plot(res)
bc$score()
```

Now we have a much more even distribution of weights and treatment/control balance.
```{r}
males_per_cage(bc)
weight_d(bc)
group_d(bc)
```
