---
title: "Batch effects and false positives: a simulation study"
author: "Jitao david Zhang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Necessity of experiment design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

In this document, we demonstrate the necessity of a proper experiment design
with a generative model which we use to simulate data with "batch" effects. We
show that a proper experiment design helps experimentalists and analysts make
correct inference about the quantity of interest that is robust against
randomness.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(designit)
library(tidyverse)
```

## A simple case study about plate effect: the background

Assume we perform an experiment to test the effect of eleven drug candidates
under development on cell viability. To do so, we treat cells in culture with a
fixed concentration of each of the eleven candidates, and we treat cells with
DMSO (dimethyl sulfoxide) as a vehicle control, since the drug candidates are
all solved in DMSO solutions. 

To assess the effect with regard to the variability intrinsic to the experiment
setup, we measure the effect of each drug candidate (and DMSO) in eight
different batches of cells, which are comparable to each other.

In total, we have 96 samples: 11 drug candidates plus one DMSO control, 8
samples each. The samples neatly fit into a 96-well microtiter plate with 8
rows, and 12 columns.

In order to avoid batch effects and to make the operation simple, all operations
and measurements are done by the same careful operator and performed at the same
time. The operator has two possibilities:

1. She does *not* randomize the samples with regard to the plate layout. The
   naive layout will put each drug candidate or control (DMSO) in one column.
   For simplicity, let us assume that the cells treated with DMSO are put in
   column 1, and cells treated with the eleven drug candidates are put in
   columns 2 to 12.
2. She randomizes the samples with regard to the plate layout, so that nearby
   samples are not necessarily of the same condition.

What is the difference between the two variants? Option 2 apparently involves
more planning and labor than option 1. If manual instead of robotic pipetting is
involved, option 2 is likely error-prone. So why bothering considering the later
option?

Randomization pays off when unwanted variance is large enough so that it may
distort our estimate of the quantity in which we are interested in. In our
example, the unwanted variance may come from a *plate effect*: due to variances
in temperature, humidity, and evaporation between wells in the plate, cells may
respond differently to *even the same treatment*. Such *plate effects* are
difficult to judge practically because they are not known prior to the
experiment, unless a calibration study is performed where the cells in a
microtiter plate are indeed treated with the same condition and measurements are
performed in order to quantify the plate effect. However, it is simple to
*simulate* such plate effects *in silico* with *a generative model*, and test
the effect of randomization.

For simplicity, we make following further assumptions:

(1) The plate effect is radial, i.e. cells in wells on the edges are more
    affected by than cells in wells in the middle of the plate. 
(2) The plate effect is positive, i.e. cells in edge wells show higher viability
    than cells in the middle wells.
(3) None of the tested compounds regulate cell viability significantly, i.e.
  cells treated with compounds and cell treated with DMSO control have the same
  expected value of viability. We simulate the effect of DMSO and compounds by
  drawing random samples from a normal distribution.
(4) The true effect of compounds and the plate effect is additive, i.e. our
    measurement is the sum of the true effect and the plate effect.

```{r}
set.seed(2307111)

conditions <- c("DMSO", sprintf("Compound%02d", 1:11))
# set up batch container
bc <- BatchContainer$new(
  dimensions = list(
    row = 8, col = 12
  )
) |>
  # assign samples with conditions and true effects
  assign_in_order(
    data.frame(
      SampleIndex = 1:96,
      Compound = factor(rep(conditions, 8), levels = conditions),
      trueEffect = rnorm(96, mean = 10, sd = 1)
    )
  )
```

## Simulating a study in which randomization is not used

First we simulate a study in which randomization is not used. In this context,
it means that the treatment (controls and compounds in columns) and the plate
effect are correlated. The following plot visualizes the layout of the plate,
the true effect, the plate effect, and the measurement as a sum of the true
effect and the plate effect.

```{r}
# get observations with batch effect
get_observations <- function(bc) {
  bc$get_samples() |>
    mutate(
      plateEffect = 0.5 * sqrt((row - 4.5)^2 + (col - 6.5)^2),
      measurement = trueEffect + plateEffect
    )
}
```

```{r}
dat <- get_observations(bc)

head(dat) |> gt::gt()
```

```{r rawPlatePlots, fig.height=5.5, fig.width=8}
cowplot::plot_grid(
  plotlist = list(
    plot_plate(dat,
      plate = plate,
      row = row, column = col, .color = Compound,
      title = "Layout by treatment"
    ),
    plot_plate(dat,
      plate = plate, row = row, column = col, .color = trueEffect,
      title = "True effect"
    ),
    plot_plate(dat,
      plate = plate, row = row, column = col, .color = plateEffect,
      title = "Plate effect"
    ),
    plot_plate(dat,
      plate = plate, row = row, column = col, .color = measurement,
      title = "Measurement"
    )
  ), ncol = 2, nrow = 2
)
```

When we perform an one-way ANOVA test with the true effect, the F-test suggests
that there are no significant differences between the treatments (p>0.05).

```{r}
summary(aov(trueEffect ~ Compound, data = dat))
```

However, if we consider the measurement, which sums the true effect and the
plate effect, the F-test suggests that there are significant differences between
the compounds (p<0.01).

```{r}
summary(aov(measurement ~ Compound, data = dat))
```

To verify, we calculate Turkey's honest significant differences using true
effect. As expected, no single compound shows significant difference from the
effect of DMSO (adjusted p-value>0.05)

```{r}
versusDMSO <- paste0(conditions[-1], "-", conditions[1])
trueDiff <- TukeyHSD(aov(
  trueEffect ~ Compound,
  data = dat
))$Compound
trueDiff[versusDMSO, ]
```

However, calculating the differences with measurements reveal that Compound 6
would have a significant difference in viability from that of DMSO (adjusted
p<0.01).

```{r}
measureDiff <- TukeyHSD(aov(measurement ~ Compound,
  data = dat
))$Compound
measureDiff[versusDMSO, ]
```

We can also detect the difference visually with a Box-Whisker plot.

```{r boxplot, fig.height=5, fig.width=5}
ggplot(
  dat,
  aes(x = Compound, y = measurement)
) +
  geom_boxplot() +
  ylab("Measurement [w/o randomization]") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


Given that our simulation study assumed that no single compound affects cell
viability significantly differently from DMSO controls. So the addition of plate
effect causes one false discovery in this simulation. It can be expected that
the false-discovery rate may vary depending on the relative strength and
variability of the plate effect with regard to the true effects. What matters
most is the observation that in the presence of plate effect, a lack of
randomization, i.e. a correlation of treatment with plate positions, may cause
wrong inferences.

## Randomization prevents plate effect from interfering with inferences

Now we use the all but one assumptions made above, with the only change that we
shall randomize the layout of the samples. The randomization will break the
correlation between treatments and plate effects.

We use the builting function `mk_plate_scoring_functions` to define the scoring
functions for the plate layout. We then use the `optimize_design` function to
randomize the layout of the samples.

```{r, eval=FALSE}
set.seed(2307111)

bc_rnd <- optimize_design(
  bc,
  scoring = mk_plate_scoring_functions(bc,
    row = "row", column = "col",
    group = "Compound"
  )
)
```

```{r, include=FALSE}
# this is quite slow, we use cached results
# bc_rnd$get_samples(include_id=TRUE) |> pull(.sample_id) |> dput()

bc_rnd <- bc$move_samples(
  location_assignment =
    c(
      1, 24, 57, 4, 91, 94, 8, 47, 27, 26, 66, 65, 53,
      67, 87, 13, 42, 60, 38, 86, 58, 21, 88, 71, 82, 18,
      56, 11, 77, 64, 31, 45, 85, 25, 3, 36, 69, 75, 50,
      96, 46, 83, 52, 89, 79, 78, 20, 92, 35, 2, 73, 32,
      16, 9, 34, 63, 54, 41, 84, 19, 90, 40, 23, 55, 61,
      29, 12, 68, 74, 39, 70, 33, 80, 5, 48, 15, 93, 49,
      30, 10, 59, 7, 14, 28, 62, 22, 43, 6, 51, 44, 81,
      72, 17, 76, 95, 37
    )
)
```

We add plate effect to the randomized data and calculate the measurement.

```{r}
dat_rnd <- get_observations(bc_rnd)

dat_rnd |>
  head() |>
  gt::gt()
```

```{r randomPlatePlots, fig.height=5.5, fig.width=8}
cowplot::plot_grid(
  plotlist = list(
    plot_plate(dat_rnd,
      plate = plate,
      row = row, column = col, .color = Compound,
      title = "Layout by treatment"
    ),
    plot_plate(dat_rnd,
      plate = plate, row = row, column = col, .color = trueEffect,
      title = "True effect"
    ),
    plot_plate(dat_rnd,
      plate = plate, row = row, column = col, .color = plateEffect,
      title = "Plate effect"
    ),
    plot_plate(dat_rnd,
      plate = plate, row = row, column = col, .color = measurement,
      title = "Measurement"
    )
  ), ncol = 2, nrow = 2
)
```

When we apply the F-test, we detect no significant differences between any
compound and DMSO.

```{r}
randMeasureDiff <- TukeyHSD(aov(measurement ~ Compound,
  data = dat_rnd
))$Compound
randMeasureDiff[versusDMSO, ]
```

We can also use the boxplot as a visual help to inspect the difference between
the treatments, to confirm that randomization prevents plate effect from
affecting the statistical inference.

```{r randBoxplot, fig.height=5, fig.width=5}
ggplot(
  dat_rnd,
  aes(x = Compound, y = measurement)
) +
  geom_boxplot() +
  ylab("Measurement [with randomization]") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Discussions and conclusions

The simple case study discussed in this vignette is an application of generative
models, which means that assuming that we know the mechanism by which the data
is generated, we can simulate the data generation process and use it for various
purposes. In our cases, we simulated a linear additive model of true effects of
compounds and control on cell viability and the plate effect induced by
positions in a microtitre plate. Using the model, we demonstrate that (1) plate
effect can impact statistical inference by introducing false positive (and in
other case, false negative) findings, and (2) a full randomization can guard
statistical inference by reducing the bias of the plate effect.

While the case study is on the margin of being overly simple, we hope that it
demonstrates the advantage of appropriate experiment design using tools like
`designit`, as well as the necessity of statistical techniques such as
randomization and blocking in drug discovery and development.
