---
title: "Moran's I"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moran's I}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(designit)
if (!requireNamespace(spdep)) {
  install.packages("spdep")
}
library(spdep)
```

We want to score layouts using [Moran's I](https://en.wikipedia.org/wiki/Moran%27s_I).

* I > 0 if the layout has clusters
* I ~ 0 for random distributions
* I < 0 for checker board like patterns

Moran’s I is defined for numeric variables. For categorical treatments, we need a one-vs-rest encoding. 
Then, we compute Moran’s I for each level and take max(|I|) or average as the final score. This should detect clustering of any condition.

## Scoring function

Like the mk_plate_scoring_function

```{r}
mk_moran_plate_scoring_functions <- function(
  batch_container,
  plate = NULL,
  row,
  column,
  group,
  neighbor = c("rook", "queen"),
  aggregate = c("max_abs", "mean_abs"),
  style = "W",
  zero_policy = TRUE
) {
  neighbor  <- match.arg(neighbor)
  aggregate <- match.arg(aggregate)

  if (!requireNamespace("spdep", quietly = TRUE)) {
    stop("Please install 'spdep' (install.packages('spdep')) to use Moran scoring.")
  }

  # helper: build a listw for the observed (non-NA) wells on one plate
  build_listw_for_plate <- function(df_plate) {
    # df_plate must have integer row/column; keep only non-NA group rows
    df_plate <- df_plate[!is.na(df_plate[[group]]), , drop = FALSE]
    if (nrow(df_plate) < 3) return(NULL)

    n_row <- max(df_plate[[row]], na.rm = TRUE)
    n_col <- max(df_plate[[column]], na.rm = TRUE)

    # cell ids in spdep::cell2nb indexing: (col-1)*n_row + row
    cells_full <- (df_plate[[column]] - 1L) * n_row + df_plate[[row]]
    keep_cells <- unique(cells_full)

    nb_full <- spdep::cell2nb(n_row, n_col, type = neighbor)

    # restrict neighbor list to kept cells, then remap to 1..m indices
    keep_set <- sort(keep_cells)
    map <- setNames(seq_along(keep_set), keep_set)

    nb_kept <- lapply(keep_set, function(cell) {
      neigh <- nb_full[[cell]]
      neigh_kept <- neigh[neigh %in% keep_set]
      unname(map[as.character(neigh_kept)])
    })

    lw <- spdep::nb2listw(nb_kept, style = style, zero.policy = zero_policy)
    list(listw = lw, df = df_plate)
  }

  # helper: compute plate score from listw + df
  score_one_plate <- function(df_plate) {
    built <- build_listw_for_plate(df_plate)
    if (is.null(built)) return(0)

    lw <- built$listw
    df <- built$df

    g <- df[[group]]
    g <- droplevels(as.factor(g))
    levels_g <- levels(g)

    # if only one level present, no meaningful clustering to penalize
    if (length(levels_g) < 2) return(0)

    I_vals <- numeric(0)

    for (lv in levels_g) {
      x <- as.numeric(g == lv)

      # Moran's I undefined if variance is 0 (all 0 or all 1)
      if (stats::var(x) == 0) {
        I <- 0
      } else {
        # fast observed statistic (no permutation test)
        I <- spdep::moran(x, lw, n = length(x), S0 = spdep::Szero(lw))$I
        if (!is.finite(I)) I <- 0
      }
      I_vals <- c(I_vals, abs(I))
    }

    if (length(I_vals) == 0) return(0)

    if (aggregate == "max_abs") max(I_vals) else mean(I_vals)
  }

  # build one scoring function per plate present in the *current* container
  samples0 <- batch_container$get_samples()

  plates <- if (!is.null(plate)) sort(unique(samples0[[plate]])) else NA

  make_scoring_fun <- function(plate_id) {
    function(bc) {
      dat <- bc$get_samples()

      if (!is.null(plate)) {
        dat <- dat[dat[[plate]] == plate_id, , drop = FALSE]
      }

      # require row/column integer-ish
      dat[[row]]    <- as.integer(dat[[row]])
      dat[[column]] <- as.integer(dat[[column]])

      score_one_plate(dat)
    }
  }

  if (!is.null(plate)) {
    out <- stats::setNames(lapply(plates, make_scoring_fun), paste0("moran_", plates))
  } else {
    out <- list(moran = make_scoring_fun(NA))
  }

  out
}
```

# Example

```{r}

```



Run function

```{r}
# bc already has samples assigned and has columns: plate,row,column,Group (example)
moran_scoring <- mk_moran_plate_scoring_functions(
  bc,
  plate  = "plate",
  row    = "row",
  column = "column",
  group  = "Group",
  neighbor = "rook",
  aggregate = "max_abs"
)

bc_opt <- optimize_design(bc, scoring = moran_scoring, max_iter = 1000)
```





You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
