---
title: "Nested dimension example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested dimension example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(designit)
library(tidyverse)
```


# Sample annotation overview

```{r}
data("multi_trt_day_samples")
```

Samples are grouped by Treatment and Collection time with the following group sizes:

```{r}
multi_trt_day_samples %>% count(Time, Treatment) %>% 
  gt::gt()
```

Total number of samples is: `r nrow(multi_trt_day_samples)`

# Task

Samples are to be blocked in batches for scRNA-seq. 

* 8 samples can be processed per day (batch)
* Within day they need to be split into 2 parallel runs (4 + 4). 

# Optimize batch

```{r}
# Setting up the batch container
bc <- BatchContainer$new(
  dimensions = c(batch = ceiling(nrow(multi_trt_day_samples) / 8), 
                 run=2, position=4)
)

# Add samples to container
bc$samples_df <- multi_trt_day_samples
# Initial random assignment
assign_in_order(bc)
# Set scoring function
bc$scoring_f <- function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score

bc
```

The samples are distributed to 4 batches (processing days).
This is done using osat scoring on sample `Treatment` and `Time`, optimizing by shuffling.

```{r}
n_shuffle <- rep(c(32, 10, 2), c(100, 80, 20)) 
n_iterations <- length(n_shuffle)

set.seed(42) # should we have conventions for this?

# initial score
#bc$score()
trace <- assign_score_optimize_shuffle(
   bc,
   n_shuffle = n_shuffle, 
   iterations = n_iterations) # default is 1000?
```

NOTE: Here the shuffling procedure is short, as it was optimized for this vignette. 
I practice you will have to run for a much higher number of iterations.

## Optimization trace 

```{r, fig.width=5, fig.height= 4}
qplot(x = 1:trace$n_steps, y = trace$scores, color = factor(c(32, n_shuffle)), 
      main = str_glue("Final score={bc$score()}"), geom = "point")
```

## Final batch layout
```{r, fig.width=6, fig.height=5}
bc$get_samples(assignment = TRUE) %>%
  mutate(batch = factor(batch)) %>% 
  ggplot(aes(x = batch, fill =  Treatment, alpha = factor(Time))) + 
  geom_bar()
```

# Repeat but use shuffle with contraints 

```{r, fig.width=6, fig.height= 4}

# clone batch container for second optimization
bc2 <- bc$clone()
# Initial random assignment
assign_in_order(bc2)

n_iterations <- 200

set.seed(42) # should we have conventions for this?

trace2 <- assign_score_optimize_shuffle(
   bc2,
   n_shuffle = NULL,
   shuffle_proposal = shuffle_with_constraints(
    src = TRUE,
    # batch needs to change for shuffle to be accepted
    dst=.src$batch != batch
   ),
   iterations = n_iterations)

qplot(x = 1:trace$n_steps, y = trace2$scores, # color = factor(n_shuffle), 
      main = str_glue("Final score={bc2$score()}"), geom = "point")

bc2$get_samples(assignment = TRUE) %>%
  mutate(batch = factor(batch)) %>% 
  ggplot(aes(x = batch, fill =  Treatment, alpha = factor(Time))) + 
  geom_bar()
```
NOTE: It is not possible to calculate the theoretically minimal osat score, right?

# Optimize runs within batch

Using shuffle with constraints

Within each day there will be 2 runs (samples processed together) with 4 samples each. 
For this we keep the optimized `batch` and now only optimize `run` with constraint.

```{r}
n_iterations <- 100

# assing new optimization function
bc$scoring_f <- function(samples) osat_score(samples, c("run"), c("Treatment", "Time"))$score
# like this the optimization score is wrong because it tries to optimize across Batches.
# Possible ways to go:
# - we'd need something like c("batch", batch/run") for optimize by batch and run within batch.
# - or we add "batch/run" to the constraints somehow.
bc$score()

trace_run <- assign_score_optimize_shuffle(
  bc,
  n_shuffle = NULL,
  shuffle_proposal = shuffle_with_constraints(
    src = TRUE,
    # batch remains the same and run needs to change
    dst = batch == .src$batch & run != .src$run
  ),
  iterations = n_iterations
)
```

```{r, fig.width=6, fig.height= 4}
qplot(x = 1:trace_run$n_steps, y = trace_run$scores, color = factor(n_iterations), 
      main = str_glue("Final score={bc$score()}"), geom = "point")
```

## Final run layout

This is not giving the expecte mix of treatments across runs.

```{r, fig.width=6, fig.height= 4}

bc$get_samples() %>% 
 mutate(run = factor(run)) %>% 
  ggplot(aes(x = run, fill =  Treatment, alpha = factor(Time))) + 
  geom_bar() +
  facet_wrap(~batch)

```


<!-- # Optimize in one go? How -->

<!-- ... from here it's not working -->

<!-- ```{r} -->
<!-- # Setting up the batch container -->
<!-- bc <- BatchContainer$new( -->
<!--   dimensions = c(batch = ceiling(nrow(multi_trt_day_samples) / 8),  -->
<!--                  run=2, position=4) -->
<!-- ) -->

<!-- # Add sample_sheet to container -->
<!-- bc$samples_df <- multi_trt_day_samples -->
<!-- # Initial random assignment -->
<!-- assign_in_order(bc) -->
<!-- ``` -->

<!-- The samples are distributed to 4 batches (processing days). -->
<!-- This is done using osat scoring on sample `Treatment` and `Time`, optimizing by shuffling. -->

<!-- ```{r} -->
<!-- n_shuffle <- rep(c(32, 10, 5, 2), c(100, 100, 100, 100))  -->
<!-- #n_shuffle <- rep(c(32, 10, 5, 2), c(50, 50, 50, 200))  -->
<!-- n_iterations <- length(n_shuffle) -->

<!-- set.seed(42) # should we have conventions for this? -->

<!-- # Set scoring function -->
<!-- bc$scoring_f <- function(samples) osat_score(samples, c("batch", "run"), c("Treatment", "Time"))$score -->

<!-- # initial score -->
<!-- #bc$score() -->
<!-- trace <- assign_score_optimize_shuffle(bc,  -->
<!--                                        n_shuffle = n_shuffle,  -->
<!--                                        iterations = n_iterations) # can we simplify? -->
<!-- # final score -->
<!-- #bc$score() -->
<!-- ``` -->

<!-- ## Optimization trace -->

<!-- ```{r, fig.width=6, fig.height= 4} -->
<!-- qplot(x = 1:trace$n_steps, y = trace$scores, color = factor(c(32, n_shuffle)),  -->
<!--       main = str_glue("Final score={bc$score()}"), geom = "point") -->
<!-- ``` -->

<!-- ## Final batch layout -->
<!-- ```{r,  fig.width=6, fig.height= 4} -->
<!-- bc$get_samples(assignment = TRUE) %>% -->
<!--   mutate(batch = factor(batch)) %>%  -->
<!--   ggplot(aes(x = batch, fill =  Treatment, alpha = factor(Time))) +  -->
<!--   geom_bar() -->

<!-- bc$get_samples(assignment = TRUE) %>%  -->
<!--  mutate(run = factor(run)) %>%  -->
<!--   ggplot(aes(x = run, fill =  Treatment, alpha = factor(Time))) +  -->
<!--   geom_bar() + -->
<!--   facet_wrap(~batch) -->
<!-- ``` -->

