---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# designit

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of designit is to ... **TODO** fill this!

## skeleton

* a metric of how balanced?
* check ^2 and ^3, it should be different (3 better)

### algorithm

- generate formula
  - .^m m = number of factors
  - check model.matrix
  - check how many to remove, and what order
  - if one order is not complete, then do random draw from that?
  - or use optMonteCarlo <-- this doesn't work
- optFederov
- optBlock
- rename Block and levels
- do it again

**should we support more than 2 levels?**

**what to do with numeric covariates**


```{r}
a <- readxl::read_excel('inst/extdata/CEA-PRIT_CEAmodels_for randomization.xlsx')
```


## Installation

TODO


```{r}
library(tidyverse)
x <- expand.grid(b = paste0('b', 1:4),
                 a = LETTERS[1:4]) %>% 
  slice(-3:-4)
```


Example dataset:

Cell: 5 levels (A, B, C, D, E)
Trt: 2 levels (N, T)
Time: 2 levels (24, 48)

2 batches:
Extract: 5 levels (1-5)
Prep: 2 levels (1-2)

all combinations, 80 expt
