---
title: "Optimizer examples"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Optimizer examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---
  
  ```{r, include = FALSE}
  knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r setup}
library(designit)
library(tidyverse)
```


# Sample annotation overview

```{r}
data("multi_trt_day_samples")
```

Samples are grouped by Treatment and Collection time with the following group sizes:

```{r}
multi_trt_day_samples %>% count(Time, Treatment) %>% 
  gt::gt()
```

Total number of samples is: `r nrow(multi_trt_day_samples)`

# Task

Samples are to be blocked in batches for scRNA-seq. 

* 8 samples can be processed per day (batch)
* Within day they need to be split into 2 parallel runs (4 + 4). 

This data set is also used in the nested dimensions example. Here, we focus on using different
methods for the optimization.

# Setting up batch container

We allocate surplus positions in the batch container and some excluded positions to check that all optimization methods support empty
container positions.

```{r}
# Setting up the batch container
bc <- BatchContainer$new(
  dimensions = c(batch = ceiling(nrow(multi_trt_day_samples) / 8), 
                 run=2, position=5),
  exclude = tibble(batch=4, run=c(1,2), position=c(5,5))
)

# Add samples to container
bc$samples_df <- multi_trt_day_samples
# Initial random assignment
assign_in_order(bc)
# Set scoring function
bc$scoring_f <- function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score

bc
```

# First optimization with fixed shuffling protocol

The samples are distributed to 4 batches (processing days).
We use the osat scoring on sample `Treatment` and `Time`, using first a shuffling protocol 
with a fixed number of sample swaps on each iteration.

Note that doing 32 swaps on 38 free container positions does not make sense, since each swapping operation affects
two different positions anyway. The upper limit is reduced to the max number of meaningful swaps (19) on the fly.

Optimization finishes after the list of permutations is exhausted.

```{r}

n_shuffle <- rep(c(32, 10, 5, 2, 1), c(100, 200, 200, 300, 300))


set.seed(42)

bc1 = bc$clone()

trace1 = optimize_design(
  bc1,
  n_shuffle = n_shuffle # will implicitly generate a shuffling function according to the provided schedule
)

trace1$elapsed

```

## Optimization trace 

Custom plot with some colours:

```{r, fig.width=5, fig.height= 4}
qplot(x = seq_along(trace1$scores), y = trace1$scores, color = factor(n_shuffle)[1:length(trace1$scores)], geom = "point") +
  labs(title="Score 1 tracing", subtitle = str_glue("Final score = {bc1$score()}"), x="Iteration",y="Score",color="n_shuffle")

```

Using the internal method...

```{r, fig.width=5, fig.height= 4}

trace1$plot()

```

We may safely apply the batch container methods get_samples() and score() also after using the new optimization code.

## Final batch layout
```{r, fig.width=6, fig.height=5}

bc1$score()

bc1$get_samples(assignment = TRUE) %>%
  mutate(batch = factor(batch)) %>% 
  ggplot(aes(x = batch, fill =  Treatment, alpha = factor(Time))) + 
  geom_bar()
```

## Perform new iterations on optimized batch container

Further optimization (using a different shuffling protocol maybe) can be done immediately
on the same batch container.

```{r}

n_shuffle <- rep(c(5, 2, 1), c(100, 100, 200))

optimize_design(
  bc1,
  n_shuffle = n_shuffle
)


```

# Optimization with specified stopping criteria 

Starting optimization from scratch, we are passing now some stopping criteria that may terminate optimization
before a shuffling protocol has been exhausted.

For demonstration, we use a shuffling function now that will do 3 sample (position) swaps per iteration
and can be called an arbitrary number of times. Thus, iteration has to be stopped by either the max_iter
criterion or by reaching a specific target score.

```{r}

bc2 = bc$clone()

trace2 = optimize_design(
  bc2,
  n_shuffle = 3, # will implicitly generate a shuffling function that will do 3 swaps at each iteration
  max_iter = 1000,
  min_score = 2.3
)

trace2$elapsed

```


# Optimization with multi-variate scoring function

Instead of passing a single scoring function, a list of multiple scoring functions can be assigned to a batch container,
each of which to return a scalar value on evaluation.

By default, the first function is determining the overall (aggregated) score for the comparison of solutions.
However, the user may specify other methods for the aggregating of the scores.

The second scoring function used here is rather redundant and just serves as illustration.

```{r}

bc3 = bc$clone()

bc3$scoring_f <- list( function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score,
                       function(samples) osat_score(samples, c("batch"), c("Treatment"))$score
)


trace3 = optimize_design(
  bc3,
  n_shuffle = 3,
  max_iter = 500,
  min_score = 2.6
)

```

Next, we repeat the optimization by using an aggregation function that takes the maximum value of the score vectors to
decide on tghe best iteration. (It has the same effect in this example though.)

```{r}

bc4 = bc$clone()

bc4$scoring_f <- list( function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score,
                       function(samples) osat_score(samples, c("batch"), c("Treatment"))$score
)


trace4 = optimize_design(
  bc4,
  n_shuffle = 3,
  aggregate_scores_func = worst_score_only,
  max_iter = 500,
  min_score = 2.6
)


```

As a final example, we calculate the (squared) L2 norm to actually aggregate the two scores. Not that this
makes sense in this case, but it could be used in future cases where optimization is carried on meaningful
distance vectors or normalized n-tuples.

We omit the `n_shuffle` parameter here, which will lead by default to pairwise sample swaps being done on each iteration.

```{r}

bc4 = bc$clone()

bc4$scoring_f <- list( function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score,
                       function(samples) osat_score(samples, c("batch"), c("Treatment"))$score
)

trace4 = optimize_design(
  bc4,
  aggregate_scores_func = L2s_norm,
  max_iter = 500,
  min_score = 2.6
)


```

# Passing a customized shuffling function

It is recommended to use the `n_shuffle` parameter to steer the optimization protocol. However, you may also provide
a dedicated shuffling function that on each call has to return a shuffling order (as integer vector) or a list with the
source and destination positions (src and dst) of the sample positions to be swapped.

The following example uses a template for creating complete random shuffles across all available positions in the batch container.
Note that this is usually not a good strategy for converging to a solution.

```{r}

bc5 = bc$clone()

trace5 = optimize_design(
  bc5,
  shuffle_proposal_func = mk_complete_random_shuffling_function(n_samples = nrow(bc4$get_samples())),
  max_iter = 500,
  min_score = 2.6
)

```

# Using simulated annealing (SA) for optimization

Esp. for very large search spaces, better solutions can be quite successfully obtained by a SA protocol which
allows the optimizer to jump over 'energy barriers' to more likely converge at lower local minima.

The optimizer usually remembers the permutation with the best overall score to start with, but this behavior
can be changed by supplying a simulated annealing protocol, most simply by generating a ready-made function template.

It is generally recommended for SA to make small changes at each step, like allowing just 1 sample swap per iteration.

```{r}

bc6 = bc$clone()

trace6 = optimize_design(
  bc6,
  n_shuffle = 1,
  acceptance_func = mk_simanneal_acceptance_func(),
  max_iter = 500
)

```
The trace may show a non strictly monotonic behavior now, reflecting the SA protocol at work.

```{r, fig.width=5, fig.height= 4}

trace6$plot()
```

Better results and quicker convergence may be achieved by playing with the starting temperature (T0) and
cooling speed (alpha) in a specific case.

```{r}

bc6 = bc$clone()

trace6 = optimize_design(
  bc6,
  n_shuffle = 1,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 100, alpha=2)),
  max_iter = 200
)

```

# Full blown example

The following example puts together all possible options to illustrate flexibility of the optimization.

```{r}

bc7 = bc$clone()

bc7$scoring_f <- list( function(samples) osat_score(samples, c("batch"), c("Treatment", "Time"))$score,
                       function(samples) osat_score(samples, c("batch"), c("Treatment"))$score,
                       function(samples) osat_score(samples, c("batch"), c("Time"))$score
)

n_shuffle <- rep(c(3, 2, 1), c(100, 100, 400))

trace7 = optimize_design(
  bc7,
  n_shuffle = n_shuffle,
  aggregate_scores_func = worst_score_only,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 500, alpha=1)),
  max_iter = 200,
  min_score = 2.3
)

trace7$plot()

bc7$get_samples(assignment = TRUE) %>%
  mutate(batch = factor(batch)) %>% 
  ggplot(aes(x = batch, fill =  Treatment, alpha = factor(Time))) + 
  geom_bar()

```

