---
title: "Sample handling order randomization"
output: 
  rmarkdown::html_vignette:
    html_document:
    df_print: paged
    mathjax: default
    number_sections: true
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Sample handling order randomization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates the use of the `desingit` when providing a
randomization scheme for sample handling order in a study, which is crucial for
minimizing bias in experimental designs.

When dealing with a large number of samples, it is often necessary to randomize
the order in which samples are handled. This can help ensure that any potential
biases related to sample handling are minimized. Bias can arise from factors
such as operator fatigue, or environmental conditions that may change over the
course of sample processing like temperature fluctuations or equipment
performance variations.

To minimize the bias we generate a sample handling order in which the
experimental conditions of interest are distributed evenly across the sample
processing procedure.

```{r include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.height = 6,
  fig.width = 6,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(designit)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)
library(ComplexHeatmap)
```

# Sample simulation

We first generate a hypothetical in-vivo dataset of samples, which includes a
treatment factor and a sex factor. The treatment factor represents different
experimental conditions.

```{r generate-data}
set.seed(42)
# Generate a hypothetical dataset
n_samples <- 40
n_conditions <- 4
# create a data frame with 40 samples with  ids, treatments and sex
samples <- data.frame(
  sample_id = 1:n_samples,
  treatment = paste0(
    "Trt",
    rep(1:n_conditions, each = ceiling(n_samples / n_conditions))[1:n_samples]
  ),
  sex = rep(c("F", "M"), n_samples / 2)
)
```

Overview of conditions

```{r}
samples |>
  count(treatment, sex) |>
  pivot_wider(names_from = sex, values_from = n) |>
  gt::gt()
```

# Designit set up

## Scoring functions

We would like to process samples in an order that is balanced by

* treatment and 
* sex

To score even distribution of samples across along the processing order we use
a $\chi^2$ test

```{r}
# scoring function for treatment
score_treatment <- function(bc) {
  x <- bc$get_samples()$treatment
  mx <- chisq.test(sort(x), sort(x))$statistic
  chisq.test(x, sort(x))$statistic / mx
}

# scoring function for sex
score_sex <- function(bc) {
  x <- bc$get_samples()$sex
  mx <- chisq.test(sort(x), sort(x))$statistic
  chisq.test(x, sort(x))$statistic / mx
}
```

## Batch container

We create a `BatchContainer` object to hold the samples and their processing
order. The `BatchContainer` allows us to manage the samples and apply
randomization and optimization functions.

```{r message=FALSE, warning=FALSE}
set.seed(42)

# make batch container with one dimension of sample processing order
bc <- BatchContainer$new(
  dimensions = list(
    `Sample processing order` = nrow(samples)
  )
)

# initial sample assignment
bc <- assign_random(bc, samples)
```

## Optimization

```{r message=FALSE, warning=FALSE}
bc <- optimize_design(
  bc,
  scoring = list(treatment = score_treatment, sex = score_sex),
  aggregate_scores_func = sum_scores,
  max_iter = 100
)
```

The score decreases quickly and reaches a stable solution

```{r}
bc$plot_trace()
```

### Visualize Sample Handling order

```{r fig.width=9}
ComplexHeatmap::HeatmapAnnotation(
  df = bc$get_samples() |>
    select(treatment, sex)
) |>
  ComplexHeatmap::plot.HeatmapAnnotation()
```

Let's say samples will be transferred into a 96-well plate 
(e.g., Sample 1 in A1, Sample 2 in A2, etc.). 
We can easily transform the sample handling order into a plate layout according to the sample handling.

```{r message=FALSE, warning=FALSE}
# generate plate layout
samples_with_order <- bc$get_samples() |>
  arrange(`Sample processing order`) |>
  mutate(
    `Well ID` = paste(rep(LETTERS[1:8], each = 12), rep(1:12, 8))[
      seq_len(nrow(bc$get_samples()))
    ]
  )
```

