---
title: "OSAT and scoring functions"
output: rmarkdown::html_vignette
params:
  iterations: 1000
vignette: >
  %\VignetteIndexEntry{OSAT and scoring functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  fig.width=6,
  fig.height=6
)
```

```{r setup}
library(designit)
library(tidyverse)
requireNamespace("OSAT")
```

Loading samples. We add two dummy columns to demonstrate how to choose batch columns of interest.
```{r}
osat_data_path <- system.file("extdata", package = "OSAT")
samples <- read_tsv(file.path(osat_data_path, "samples.txt"),
  col_types = cols(SampleType = col_factor(), Race = col_factor(), AgeGrp = col_factor())
) %>%
  mutate(dummy_var1 = rnorm(n()), dummy_var2 = str_c(SampleType, Race, sep = " "))
```

# Running OSAT optimization
Here we use OSAT to optimize setup.
```{r}
gs <- OSAT::setup.sample(samples, optimal = c("SampleType", "Race", "AgeGrp"))

gc <- OSAT::setup.container(OSAT::IlluminaBeadChip96Plate, 7, batch = "plates")

set.seed(1234)

bench::system_time(
  g_setup <- OSAT::create.optimized.setup(sample = gs, container = gc, nSim = params$iterations)
)

OSAT::QC(g_setup)
```

Saving starting point of optimization
```{r}
set.seed(1234)

g_setup_start <- OSAT::create.optimized.setup(sample = gs, container = gc, nSim = 1) %>%
  OSAT::get.experiment.setup()
```


Visualize various batch factors. OSAT score is optimized only for `plates` in this case.
```{r}
OSAT::get.experiment.setup(g_setup) %>%
  select(AgeGrp, plates, chipRows, chipColumns, chips, rows, columns, wells) %>%
  pivot_longer(-AgeGrp) %>%
  count(AgeGrp, value, name) %>%
  ggplot(aes(AgeGrp, n, fill = factor(value))) +
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free_y")
```

## Visualize for plates
```{r}
plot_batch <- function(df) {
  df %>%
    select(plates, SampleType, Race, AgeGrp) %>%
    pivot_longer(c(SampleType, Race, AgeGrp), names_to = "variable", values_to = "level") %>%
    count(plates, variable, level) %>%
    ggplot(aes(level, n, fill = factor(plates))) +
    geom_col(position = "dodge") +
    facet_wrap(~variable, scales = "free", ncol=1)
}
```

Before the optimization.
```{r}
g_setup_start %>% plot_batch()
```

After the optimization.
```{r}
OSAT::get.experiment.setup(g_setup) %>%
  plot_batch()
```


# Compare scores with various implementations
Compare OSAT score generated using various methods.
```{r}
# scores using datatable implementation
OSAT::getLayout(gc) %>%
  left_join(OSAT::get.experiment.setup(g_setup)) %>%
  data.table::data.table() %>%
  osat_score("plates", c("SampleType", "Race", "AgeGrp")) %>%
  .$score

# scores using slower tibble implementation
OSAT::getLayout(gc) %>%
  left_join(OSAT::get.experiment.setup(g_setup)) %>%
  designit:::osat_score_tibble(plates, c(SampleType, Race, AgeGrp)) %>%
  .$score

# score using OSAT
g_setup@metadata$optValue %>% tail(1)
```


# Run using BatchContainer
First let's create a BatchContainer with same dimensions.
```{r}
bc <- BatchContainer$new(
  dimensions = c(plates = 7, chips = 8, rows = 6, columns = 2)
)
bc

bc$n_available
```

Assign samples and get initial setup.
```{r}
bc$samples_df <- samples

starting_assignment <- bc$locations_df %>%
  left_join(g_setup_start) %>%
  pull(ID) %>%
  as.integer()

bc$assignment_vec <- starting_assignment

bc$get_samples(remove_empty_locations = TRUE) %>%
  plot_batch()
```

## Using designit OSAT score implementation
```{r}
bc$scoring_f <- function(samples) osat_score(samples, "plates", c("SampleType", "Race", "AgeGrp"))$score

bc$score()
g_setup@metadata$optValue %>% head(1)
# should be identical

bench::system_time({
  set.seed(123)
  trace <- assign_score_optimize_shuffle(bc, iterations = params$iterations)
})
```

```{r}
# final score
bc$score()
plot(trace, main = str_glue("Final score={bc$score()}"))
bc$get_samples(remove_empty_locations = TRUE) %>%
  plot_batch()
```

## Cache `expected_dt`
Caching `expected_dt` should improve the performance.

```{r}
bc$assignment_vec <- starting_assignment

# precompute expected_dt
edf <- osat_score(bc$samples_dt, "plates", c("SampleType", "Race", "AgeGrp"))$expected_dt

bc$scoring_f <- function(samples) osat_score(samples, "plates", c("SampleType", "Race", "AgeGrp"), expected_dt = edf)$score

bench::system_time({
  set.seed(123)
  trace <- assign_score_optimize_shuffle(bc, iterations = params$iterations)
})
```

## Manually work with `data.table`
Instead of relying on `BatchContainer`, here we have a manual optimization process using `data.table`.

```{r}
bc$assignment_vec <- starting_assignment

fast_osat_optimize <- function(bc, batch_vars, feature_vars, iterations) {
  ldf <- data.table::data.table(bc$locations_df)[, c("plates")][, ".sample_id" := bc$assignment_vec]
  fcols <- c(".sample_id", feature_vars)
  smp <- data.table::data.table(bc$samples_df)[, ..fcols]
  df <- smp[ldf, on = ".sample_id"]

  v <- osat_score(df, batch_vars, feature_vars)
  edf <- v$expected_dt
  current_score <- v$score
  scores <- numeric(length = iterations)
  n_avail <- nrow(df)

  for (i in 1:iterations) {
    repeat {
      pos <- sample(n_avail, 2)

      # does not make sense to shuffle NAs
      if (any(!is.na(df[pos, feature_vars[1]]))) {
        break
      }
    }

    val <- df[c(pos[2], pos[1]), fcols, with = FALSE]
    df[c(pos[1], pos[2]), (fcols) := val]

    new_score <- osat_score(df, batch_vars, feature_vars, edf)$score
    if (new_score <= current_score) {
      current_score <- new_score
    } else {
      df[c(pos[2], pos[1]), (fcols) := val]
    }

    scores[i] <- current_score
  }

  bc$assignment_vec <- df$.sample_id

  scores
}

bench::system_time({
  set.seed(123)
  trace <- fast_osat_optimize(bc, "plates", c("SampleType", "Race", "AgeGrp"), iterations = params$iterations)
})
```


# Score demonstration
```{r}
bc$score()
bc$scoring_f <- function(...) rnorm(1)
bc$score()
```

```{r}
assign_random(bc)

bc$get_samples()
bc$get_samples(remove_empty_locations = TRUE)

bc$score()
bc$score(aux = TRUE)

bc$aux_scoring_f <- list(
  function(samples) rnorm(1, 100),
  function(samples) -7
)

bc$scoring_f <- function(samples) rnorm(1) + 2 * rexp(1)

bc$aux_scoring_f <- list(
  fc1 = function(samples) rnorm(1, 100),
  fc2 = function(samples) -7
)

bc$score(aux = TRUE)
```
