---
title: "Plate scoring examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plate scoring examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(designit)
if (!requireNamespace("tidyverse", quietly = TRUE)) rlang::abort("Tidyverse required!")
```


# Example 1: An expensive way to construct a 4x4 latin square (one plate)

(latin square should give the best score)

First using a combination of two OSAT scores (for row and column)

This usually produces a latin square when using the squared L2 norm (L2s) for aggregation of the 2 scores.

```{r}
# Setting up the batch container
example1 <- BatchContainer$new(
  dimensions = c(
    plate = 1,
    row = 4, col = 4
  )
)

# Add samples to container
# Need unique Sample ID. Can we drop this constraint?
assign_in_order(example1, 
                samples = tibble::tibble(Group = rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 4), 
                                         ID = 1:16))

# The following does not work (constant score of 12!) - why?
# example1$scoring_f <- osat_score_generator(batch_vars = c("row","col"), feature_vars = c("Group"))

example1$scoring_f <- function(samples) osat_score(samples, batch_vars = c("row", "col"), 
                                                   feature_vars = c("Group"))$score

bc <- example1$copy()

# # Set scoring function
bc$scoring_f <- list(
  osat_score_generator(batch_vars = c("row"), feature_vars = c("Group")),
  osat_score_generator(batch_vars = c("col"), feature_vars = c("Group"))
)

start_time <- Sys.time()
optimize_design(bc,
  min_score = 0, max_iter = 1e4,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
Sys.time() - start_time

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```

Now using a dedicated scoring for the group distances on a plate. 

This should reliably lead to a nice symmetry-bearing latin square design with only a one-dimensional score to look at.

```{r}
bc <- example1$copy()

bc$scoring_f <- mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group")

start_time <- Sys.time()
optimize_design(bc,
  min_score = 6.1, max_iter = 1e4,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5))
)

Sys.time() - start_time

bc$score()

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```
# Example 2: Scoring two plates at once

(latin square for each plate should give the best score)
 
We set up in one go 2 plate scoring functions, each one acting locally on a specific plate, 
plus one osat score to guarantee uniform distribution of groups across plates.

The initial sample allocation (by assign_in_order) leads to a poor starting point 
since each plate has only 2 of the 4 groups represented. 

This is not a problem as long as we make sure that initial permutations are likely to remedy the situation. 
That's why we ensure 10 pairwise sample swaps for the first iterations.
 
```{r}
# Setting up the batch container
example2 <- BatchContainer$new(
  dimensions = c(
    plate = 2,
    row = 4, col = 4
  )
)

# Add samples to container
assign_in_order(example2, samples = tibble::tibble(Group = c(rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 8)),
                                                   Group_num = c(rep(1:4, each = 7), NA,NA,NA,NA), # possibility for debugging
                                                   ID = 1:32))

bc <- example2$copy()

bc$scoring_f <- c(mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group"),
  osat_plate = osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
)

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)

bc$score()

start_time <- Sys.time()
optimize_design(bc,
  n_shuffle = c(rep(10, 10), rep(3, 90), rep(2, 100), rep(1, 1800)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
  aggregate_scores_func = worst_score_only
)
Sys.time() - start_time

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```

While this 'global' optimization is possible, it does probably not converge to an (almost) ideal solution 
in an acceptable time if there are more samples involved.
This is due to a lot of unproductive sample swapping happening across the plates.

One way to address this: we may split the optimization into two cycles, 
first assigning samples to plates (balancing groups),
then improving the positions of the samples within each plate. 
This motivates the use of a dedicated sample permutation function
which takes the plate structure into account and only shuffles samples around within one plate.


```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))

start_time <- Sys.time()
optimize_design(bc,
  quiet = T,
  min_score = 2, max_iter = 5e3,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

optimize_design(bc,
  max_iter = 5e3,
  shuffle_proposal_func = mk_subgroup_shuffling_function(bc, subgroup_vars = c("plate")),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
Sys.time() - start_time

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```
In this case, the shuffling function exchanges 1 pair of sample assignments every time (the default). 
However, any number of constant swaps or a swapping protocol (formally a vector of integers) 
can be supplied as well.


Now for the most efficient solution:
we start again by first assigning samples to plates (balancing groups),
then making use of the independence of the two within-plate optimizations and improving them
one after the other. 

This is possible by passing the argument \code{restrain_on_subgroup_levels} to the function that generates
the permutations. It enforces permutation only to happen first within plate 1, then within plate 2, so that
the two scores can be optimized in succeeding runs.

```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))

start_time <- Sys.time()
optimize_design(bc,
  quiet = T,
  min_score = 2, max_iter = 5e3,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

bc$score()

optimize_design(bc,
  max_iter = 5e3,
  quiet=T,
  shuffle_proposal_func = mk_subgroup_shuffling_function(bc, subgroup_vars = c("plate"), restrain_on_subgroup_levels=c(1)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)

bc$score()

optimize_design(bc,
  max_iter = 5e3,
  quiet=T,
  shuffle_proposal_func = mk_subgroup_shuffling_function(bc, subgroup_vars = c("plate"), restrain_on_subgroup_levels=c(2)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)

bc$score()

Sys.time() - start_time

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```


# Example 3: 3 plates with different dimension and different sample group sizes

We simulate one ordinary 96 well plate and two smaller ones with sizes 6x8 and 4x6, respectively. 
There are 3 experimental groups as well with sample sizes of 69, 30 and 69, respectively. 
This example should demonstrate that an empirically determined normalization of the scores 
yield 3 comparable numerical values, independent of the different plate and group sizes.  

Again, a first optimization aims to achieve same group allocation on each plate, 
while the second run takes care of the sample distribution on each plate.

We use assign_random in this example to start from a more balanced initial position as compared to example 2.

Aggregation of scores is done by the L2s method (square of L2 norm). 
Because of the comparable numerical range of scores, 
also the worst_score method could be used for aggregation. 
However, the L2s method always takes into account all individual scores,
providing more stability in case the 3 plate scores are not exactly normalized. 

```{r}
# Setting up the batch container

example3 <- BatchContainer$new(
  dimensions = c(
    plate = 3,
    row = 8, col = 12
  ),
  exclude = dplyr::bind_rows(
    tidyr::crossing(plate = 2, row = 1:8, col = 1:12) %>% dplyr::filter(row > 6 | col > 8),
    tidyr::crossing(plate = 3, row = 1:8, col = 1:12) %>% dplyr::filter(row > 4 | col > 6)
  )
)


# Assign samples randomly to start from a better initial state
assign_random(example3, 
              samples = tibble::tibble(Group = rep.int(c("Grp 1", "Grp 2", "Grp3"), 
                                                       times = c(69, 30, 69)), 
                                       ID = 1:168))

bc <- example3$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))

start_time <- Sys.time()
optimize_design(bc,
  quiet = T,
  min_score = 1.49, max_iter = 5e3,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
)

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", 
                                           column = "col", group = "Group")

bc$score()

optimize_design(bc,
  min_score = 0, max_iter = 1e4,
  shuffle_proposal_func = mk_subgroup_shuffling_function(bc,
    subgroup_vars = c("plate"),
    n_swaps = c(rep(5, 500), rep(3, 1500), rep(2, 3000), rep(1, 5000))
  ),
  # acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 1)),
  aggregate_scores_func = L2s_norm
)

Sys.time() - start_time

plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group)
```

# Example 4: More than one group factor to balance and empty plate positions

In this example, we have 2 factors to distribute across one plate: Treatment and (animal) sex.

To indicate that the balancing of treatment is considered more important than the animal sex we assign a custom aggregation
function giving more weight to the treatment variable. 

There can be less samples than possible positions on the plate(s). In this case, we simulate 20 animal derived samples distributed
on a plate with 24 locations.


```{r}
# Setting up the batch container

example4 <- BatchContainer$new(
  dimensions = c(
    plate = 1, row = 6, col = 4
  )
)


# Assign samples randomly to start from lower score (avoid Inf values even since plate 3 will miss 2 groups initially :)
assign_in_order(example4, samples = tibble::tibble(
  Group = rep.int(c("Treatment 1", "Treatment 2"), times = c(10, 10)),
  Sex = c(rep(c("M", "F", "F", "M"), times = 4), "M",NA,NA,"F"), ID = 1:20
))

bc <- example4$copy()

cowplot::plot_grid(
  plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group),
  plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Sex),
  ncol = 2
)


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)

bc$score()

optimize_design(bc,
  min_score = 17.66, max_iter = 5e3,
  n_shuffle = 1,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 1)),
  aggregate_scores_func = function(scores, ...) {
    2 * scores["Group.Plate"] + scores["Sex.Plate"]
  }
)

Sys.time() - start_time

cowplot::plot_grid(
  plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Group),
  plot_plate(bc$get_samples(), plate = plate, row = row, column = col, .color = Sex),
  ncol = 2
)
```
