---
title: "Plate scoring examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plate scoring examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
params:
  regenerate_cache: FALSE
  verbose_cache: TRUE
---
  
  
```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(designit)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Example 1: An expensive way to construct a 4x4 latin square (one plate)

(latin square should give the best score)

First using a combination of two OSAT scores (for row and column)

This usually produces a latin square when using the squared L2 norm (L2s) for
aggregation of the 2 scores.

```{r}
# Setting up the batch container
example1 <- BatchContainer$new(
  dimensions = c(
    plate = 1,
    row = 4, col = 4
  )
)

# Add samples to container
# Need unique Sample ID. Can we drop this constraint?
assign_in_order(example1,
  samples = tibble::tibble(
    Group = rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 4),
    ID = 1:16
  )
)

# The following does not work (an gives a constant score of 144!)
# example1$scoring_f <- osat_score_generator(batch_vars = c("row","col"), feature_vars = c("Group"))
# First analysis of problem indicates that osat_score generates a full row*col vector of 'ideal scores'
# which are in fact the same value, implying an identical overall result as each position can be either
# allocated by 1 sample or 0 samples, the sum of 1's being the sample count.
# --> don't use osat_score if there's a lack of samples as compared to possible positioning

bc <- example1$copy()

# # Set scoring function
bc$scoring_f <- list(
  Row.Score = osat_score_generator(batch_vars = c("row"), feature_vars = c("Group")),
  Column.Score = osat_score_generator(batch_vars = c("col"), feature_vars = c("Group"))
)
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace1")
```

```{r eval=.eval_cached_chunk}

set.seed(41)

trace <- optimize_design(bc,
  max_iter = 300, # this is set to shorten vignette run-time based on known random seed, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 1: Using OSAT scores for plate design\n(not the recommended way!)"
)
```

Now using a dedicated scoring for the group distances on a plate. 

This should reliably lead to a nice symmetry-bearing latin square design with
only a one-dimensional score to look at.

```{r}
bc <- example1$copy()

bc$scoring_f <- mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group")
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace2")
```

```{r eval=.eval_cached_chunk}

set.seed(42)
trace <- optimize_design(bc,
  max_iter = 1000, # this is set to shorten vignette run-time based on random seed, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  quiet = TRUE
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace$elapsed

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 1: Using a dedicated plate scoring function:\nThis should show a latin square!"
)
```

# Example 2: Scoring two plates at once

(latin square for each plate should give the best score)

We set up in one go 2 plate scoring functions, each one acting locally on a
specific plate, plus one osat score to guarantee uniform distribution of groups
across plates.

The initial sample allocation (by assign_in_order) leads to a poor starting
point since each plate has only 2 of the 4 groups represented.

This is not a problem as long as we make sure that initial permutations are
likely to remedy the situation. That's why we ensure 10 pairwise sample swaps
for the first iterations.

```{r}
# Setting up the batch container
example2 <- BatchContainer$new(
  dimensions = c(
    plate = 2,
    row = 4, col = 4
  )
)

# Add samples to container
assign_in_order(example2, samples = tibble::tibble(
  Group = c(rep(c("Grp 1", "Grp 2", "Grp 3", "Grp 4"), each = 8)),
  ID = 1:32
))

bc <- example2$copy()

bc$scoring_f <- c(mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group"),
  osat_plate = osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
)

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: Initial sample arrangement"
)

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace3")
```

```{r eval=.eval_cached_chunk}
set.seed(41)
trace <- optimize_design(bc,
  n_shuffle = c(rep(10, 10), rep(3, 90), rep(2, 100), rep(1, 1400)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
  aggregate_scores_func = worst_score,
  quiet = TRUE
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace$elapsed

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: Design created by swapping samples 'globally' across the plates"
)
```

While this 'global' optimization is possible, it does probably not converge to
an (almost) ideal solution in an acceptable time if there are more samples
involved. This is due to a lot of unproductive sample swapping happening across
the plates.

One way to address this: we may split the optimization into two cycles, first
assigning samples to plates (balancing groups), then improving the positions of
the samples within each plate. This motivates the use of a dedicated sample
permutation function which takes the plate structure into account and only
shuffles samples around within one plate.

```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))

set.seed(42)
optimize_design(bc,
  quiet = TRUE,
  max_iter = 200, # this is set to shorten vignette run-time, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: 'Plate wise' design\nStep 1: after allocating samples to plates"
)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace4")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 400,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate")),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm,
  quiet = TRUE
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: 'Plate wise' design\nStep 2: after arranging samples within plates"
)
```

In this case, the shuffling function exchanges 1 pair of sample assignments
every time (the default). However, any number of constant swaps or a swapping
protocol (formally a vector of integers) can be supplied as well.


Now for the most efficient solution: we start again by first assigning samples
to plates (balancing groups), then making use of the independence of the two
within-plate optimizations and improving them one after the other.

This is possible by passing the argument \code{restrain_on_subgroup_levels} to
the function that generates the permutations. It enforces permutation only to
happen first within plate 1, then within plate 2, so that the two scores can be
optimized in succeeding runs.

```{r}
# Setting up the batch container

bc <- example2$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace5")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  quiet = TRUE,
  max_iter = 150, # this is set to shorten vignette run-time, normally we don't know.
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 0.5)),
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: 'Serial plate' design\nStep 1: after allocating samples to plates"
)

bc$scoring_f <- mk_plate_scoring_functions(bc, plate = "plate", row = "row", column = "col", group = "Group")

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace6")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 150,
  quiet = TRUE,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate"), restrain_on_subgroup_levels = c(1)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace7")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 550,
  quiet = TRUE,
  shuffle_proposal_func = mk_subgroup_shuffling_function(subgroup_vars = c("plate"), restrain_on_subgroup_levels = c(2)),
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
  aggregate_scores_func = L2s_norm
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace

bc$score()

plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 2: 'Serial plate' design\nStep 2: after optimizing each plate in turn"
)
```


# Example 3: 3 plates with different dimension and different sample group sizes

We simulate one ordinary 96 well plate and two smaller ones with sizes 6x8 and
4x6, respectively. There are 3 experimental groups as well with sample sizes of
69, 30 and 69, respectively. This example should demonstrate that an empirically
determined normalization of the scores yield 3 comparable numerical values,
independent of the different plate and group sizes.

Again, a first optimization aims to achieve same group allocation on each plate,
while the second run takes care of the sample distribution on each plate.

We use assign_random in this example to start from a more balanced initial
position as compared to example 2.

Aggregation of scores is done by the L2s method (square of L2 norm). Because of
the comparable numerical range of scores, also the worst_score method could be
used for aggregation. However, the L2s method always takes into account all
individual scores, providing more stability in case the 3 plate scores are not
exactly normalized.

```{r}
# Setting up the batch container

example3 <- BatchContainer$new(
  dimensions = c(
    plate = 3,
    row = 8, col = 12
  ),
  exclude = dplyr::bind_rows(
    tidyr::crossing(plate = 2, row = 1:8, col = 1:12) %>% dplyr::filter(row > 6 | col > 8),
    tidyr::crossing(plate = 3, row = 1:8, col = 1:12) %>% dplyr::filter(row > 4 | col > 6)
  )
)


# Assign samples randomly to start from a better initial state
assign_random(example3,
  samples = tibble::tibble(
    Group = rep.int(c("Grp 1", "Grp 2", "Grp3"),
      times = c(69, 30, 69)
    ),
    ID = 1:168
  )
)

bc <- example3$copy()

bc$scoring_f <- osat_score_generator(batch_vars = c("plate"), feature_vars = c("Group"))
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace8")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  quiet = TRUE,
  max_iter = 150,
  n_shuffle = 2,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 1000, alpha = 0.5)),
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace
```

```{r, fig.width=7, fig.height=3.5}
plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 3: Dealing with plates of different size\nStep 1: after distributing groups across plates"
)
```

```{r}
bc$scoring_f <- mk_plate_scoring_functions(bc,
  plate = "plate", row = "row",
  column = "col", group = "Group"
)

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace9")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 300,
  shuffle_proposal_func = mk_subgroup_shuffling_function(
    subgroup_vars = c("plate"),
    n_swaps = c(rep(5, 500), rep(3, 1500), rep(2, 3000), rep(1, 5000))
  ),
  # acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 1)),
  aggregate_scores_func = L2s_norm,
  quiet = TRUE
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r}
trace$elapsed

bc$score()
```

```{r, fig.width=7, fig.height=3.5}
plot_plate(bc,
  plate = plate, row = row, column = col, .color = Group,
  title = "Exa 3: Dealing with plates of different size\nStep 2: after swapping samples within plates"
)
```

# Example 4: More than one group factor to balance and empty plate positions

In this example, we have 2 factors to distribute across one plate: Treatment and
(animal) sex.

To indicate that the balancing of treatment is considered more important than
the animal sex we assign a custom aggregation function giving more weight to the
treatment variable. (A better aggregation mechanism has to be implemented!!!)

There can be less samples than possible positions on the plate(s). In this case,
we simulate 20 animal derived samples distributed on a plate with 24 locations.


```{r, fig.width=7, fig.height=3.5}
# Setting up the batch container
example4 <- BatchContainer$new(
  dimensions = c(
    plate = 1, row = 6, col = 4
  )
)


# Assign samples randomly to start from lower score (avoid Inf values even since plate 3 will miss 2 groups initially :)
assign_in_order(example4, samples = tibble::tibble(
  Group = rep.int(c("Treatment 1", "Treatment 2"), times = c(10, 10)),
  Sex = c(rep(c("M", "F", "F", "M"), times = 4), "M", NA, NA, "F"), ID = 1:20
))

bc <- example4$copy()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Initial layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Initial layout by Sex"),
  ncol = 2
)


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)

bc$score()
```

```{r include=FALSE, message = FALSE}
designit::.load_cache("trace10")
```

```{r eval=.eval_cached_chunk}
set.seed(42)
trace <- optimize_design(bc,
  max_iter = 750,
  n_shuffle = 1,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 1)),
  aggregate_scores_func = function(scores, ...) {
    2 * scores["Group.Plate"] + scores["Sex.Plate"]
  },
  quiet = TRUE
)
```

```{r include=FALSE, message = FALSE}
designit::.save_cache(trace, bc)
```

```{r, fig.width=7, fig.height=3.5}
trace$elapsed

bc$score()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)
```


We do the same example with auto-scaling, weighted scoring and SA to have a
reference!

```{r eval=.eval_cached_chunk}
bc <- example4$copy()


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)

bc$score()
cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)

set.seed(42)
trace <- optimize_design(bc,
  max_iter = 500,
  n_shuffle = 1,
  acceptance_func = mk_simanneal_acceptance_func(mk_simanneal_temp_func(T0 = 10000, alpha = 1)),
  aggregate_scores_func = function(scores, ...) {
    setNames(2 * scores["Group.Plate"] + scores["Sex.Plate"], nm = "Weighted.Score")
  },
  autoscale_scores = T,
  quiet = F
)

bc$score()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)
```

We do the  same example with auto-scaling and position-dependent scoring now,
not aggregating the score vector! This is more effective even when using the
default acceptance function. We are strictly prioritizing the leftmost score in addition
to reflect relevance for the design.

```{r eval=.eval_cached_chunk}

bc <- example4$copy()


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)

bc$score()
cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)

set.seed(42)
trace <- optimize_design(bc,
  max_iter = 5000,
  n_shuffle = 1,
  acceptance_func = accept_leftmost_improvement,
  autoscale_scores = TRUE,
  quiet = F
)

bc$score()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)
```

Using a tolerance value to accept slightly worse solutions in the leftmost relevant score if overcompensated by other scores:

```{r eval=.eval_cached_chunk}

bc <- example4$copy()


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)


set.seed(42)
trace <- optimize_design(bc,
  max_iter = 5000,
  n_shuffle = 1,
  acceptance_func = mk_accept_leftmost_improvement_with_tolerance(tolerance = 0.1),
  autoscale_scores = TRUE,
  quiet = F
)

bc$score()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)
```


Testing an alternative left-to-right weighing of scores, based on exponential
down-weighing of the respective score differences at position p with factor
kappa^p, 0<kappa<1 We choose a kappa of 0.5, i.e. the second score's improvement
counts half of that of the first one.

```{r eval=.eval_cached_chunk}
bc <- example4$copy()


bc$scoring_f <- c(
  Group = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Group"),
  Sex = mk_plate_scoring_functions(bc, row = "row", column = "col", group = "Sex")
)

bc$score()
cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)

set.seed(42)
trace <- optimize_design(bc,
  max_iter = 1000,
  n_shuffle = 1,
  acceptance_func = mk_exponentially_weighted_acceptance_func(kappa = 0.5, simulated_annealing = T),
  autoscale_scores = TRUE,
  quiet = F
)

bc$score()

cowplot::plot_grid(
  plot_plate(bc, plate = plate, row = row, column = col, .color = Group, title = "Final layout by Group"),
  plot_plate(bc, plate = plate, row = row, column = col, .color = Sex, title = "Final layout by Sex"),
  ncol = 2
)
```
